---
title: "Lab 02 - Plastic waste"
author: "John T. Bennett"
date: "Feb. 9, 2022"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
View(plastic_waste)
help()
```

Warm up:

1) R Studio panes: On right side, Source is on top of Console. On left side, 
Environment is on top of Files/Plots

2) There are 240 observations

3) NA - missing data?

## Exercises

Trinidad and Tobago stands out, which makes sense based on their lax waste 
management practices. 

```{r plastic-waste-continent}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
```

### Exercise 1.1. Plot, using histograms, the distribution of plastic waste per capita faceted by continent.

```{r plastic-waste-faceted}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap, fill = continent)) +
  geom_histogram(binwidth = 0.2) +
 labs(
   x = "Per capita plastic waste",
   title = "Distribution of plastic waste per capita, by continent"
 ) +
 facet_wrap(~ continent, nrow = 2)
```

1.1.a. Visualizing numerical data using density plots.
```{r plastic-waste-density}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_density()
```

1.1.b. Comparing distributions across continents by coloring density curves by continent. 
```{r plastic-waste-colorcont}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent)) +
  geom_density()
```

1.1.c. Filling curves with colors

```{r plastic-waste-curvecolor}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density()
```

### Exercise 2.1. Recreating density plots using lower alpha level

```{r plastic-waste-loweralpha}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density(alpha = 0.1)
```

### Exercise 2.2

I think we defined the alpha level as a characteristic of the plotting geom because this definition refers specifically to the density or transparency of what we mapped- meaning the curves.  

### Exercise 3.1. Convert side-by-side box plots to violin plots. 

Box plots:

```{r plastic-waste-box}
ggplot(data = plastic_waste, 
       mapping = aes(x = continent, 
                     y = plastic_waste_per_cap)) +
  geom_boxplot()
```
Violin plots:

```{r plastic-waste-violin}
ggplot(plastic_waste, aes(x = continent, y = plastic_waste_per_cap)) +
  geom_violin()
```

3.1.a. What do the violin plots reveal that box plots do not? 
Instead of using bars, the violin plots show us the shape of the data distribution. 

3.1.b. What features are apparent in the box plots but not in the violin plots?
The boxes show us a tidy visualization of where 50% of the data lay in the distribution. 

### Exercise 4.1. Scatterplot- relationship between plastic waste per capita and mismanaged plastic waste per capita 

```{r plastic-waste-mismanaged}
ggplot(plastic_waste, aes(x = mismanaged_plastic_waste_per_cap, y = plastic_waste_per_cap)) +
  geom_point()
```

Describe the relationship: 
Something other than plastic waste per capita must be driving mismanaged plastic waste per capita. The lowest levels of mismanaged waste and the highest levels of mismanaged waste occur within the same narrow range of plastic waste per capita. Thus, mismanaged plastic waste per capita is weakly related to plastic waste per capita. 

### Exercise 4.2. Color the points in the scatterplot by continent. 

```{r plastic-waste-mismanaged-continent}
ggplot(plastic_waste, aes(x = mismanaged_plastic_waste_per_cap, y = plastic_waste_per_cap, color = continent)) +
  geom_point()
```

4.2.a. Does there seem to be any clear distinctions between continents with respect to how plastic waste per capita and mismanaged plastic waste per capita are associated?
The noteworthy pattern, in my opinion, is that Europe has very low levels of both plastic waste per capita and mismanaged plastic waste per capita.

### Exercise 4.3. Visualize the relationship between plastic waste per capita and total population as well as plastic waste per capita and coastal population. 

4.3.a. Relationship between plastic waste per capita and total population
```{r plastic-waste-population-total}
ggplot(plastic_waste, aes(x = total_pop, y = plastic_waste_per_cap)) +
  geom_point()
```

4.3.b. Relationship between plastic waste per capita and coastal population

```{r plastic-waste-population-coastal}
ggplot(plastic_waste, aes(x = coastal_pop, y = plastic_waste_per_cap)) +
  geom_point()
```

4.3.c. Do either of these pairs of variables appear to be more strongly linearly associated?

Population appears to have a stronger relationship with plastic waste per capita than coastal population has with plastic waste per capita. 
However, appearances could be decieving. The coastal population is smaller than the total population, so the data appear more densely bunched together in the coastal population plot. In contrast, total population is spread out across a higher value on the x-axis than coastal population, making the relationships appear more diffuse. 
We need to compare the two plots more directly by comparing the two plots when the x axis is of equal length. (Or plotting both pairs of variables together at the same time) 

### Exercise 5.1.
I gave this one a shot, but nothing doing this time...

{r smooth}
ggplot(plastic_waste, aes(x = coastal_pop, y = plastic_waste_per_cap, color = continent))
+ geom_point(aes(color = continent))
  + geom_point(alpha = (1/3), size = 3) + geom_smooth(lwd = 3, se = FALSE)
  + geom_point(alpha = (1/3), size = 3) + facet_wrap(~ continent) +
  geom_smooth(lwd = 1.5, se = FALSE)

### Pro-Tips

Radioactive manta rays!

```{r protip}
ggplot(data = plastic_waste, 
       mapping = aes(x = continent, 
                     y = plastic_waste_per_cap)) +
  geom_violin()+
  geom_boxplot(width=.3, fill="green") +
  stat_summary(fun.y=median, geom="point") 
```
  


Helpful reference:http://www.sthda.com/english/wiki/ggplot2-themes-and-background-colors-the-3-elements



